A matrix-free approach for finite-strain hyperelastic problems using geometric multigrid

### 2019-08-12 doc/large_strain.tex
remove the comment on coarse solver
minor TeX improvements

\usepackage{todonotes}
\usepackage{bm}
\usepackage{xcolor}
\usepackage{listings}
\lstset{ %
   backgroundcolor=\color{white},   % choose the background color; you must add \usepackage{color} or \usepackage{xcolor}
   basicstyle=\small,        % the size of the fonts that are used for the code
   breakatwhitespace=false,         % sets if automatic breaks should only happen at whitespace
   breaklines=false,                 % sets automatic line breaking
   captionpos=b,                    % sets the caption-position to bottom
   %commentstyle=\color{mygreen},    % comment style
   deletekeywords={...},            % if you want to delete keywords from the given language
   escapeinside={\%*}{*)},          % if you want to add LaTeX within your code
   extendedchars=false,              % lets you use non-ASCII characters; for 8-bits encodings only, does not work with UTF-8
   frame=single,	                   % adds a frame around the code
   keepspaces=true,                 % keeps spaces in text, useful for keeping indentation of code (possibly needs columns=flexible)
   keywordstyle=\color{blue},       % keyword style
   language=C++,                 % the language of the code
   otherkeywords={*,...},           % if you want to add more keywords to the set
   numbers=left,                    % where to put the line-numbers; possible values are (none, left, right)
   numbersep=5pt,                   % how far the line-numbers are from the code
   numberstyle=\tiny\color{gray}, % the style that is used for the line-numbers
   rulecolor=\color{black},         % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. comments (green here))
   showspaces=false,                % show spaces everywhere adding particular underscores; it overrides 'showstringspaces'
   showstringspaces=false,          % underline spaces within strings only
   showtabs=false,                  % show tabs within strings adding particular underscores
   stepnumber=2,                    % the step between two line-numbers. If it is 1, each line will be numbered
%   stringstyle=\color{mymauve},     % string literal style
   tabsize=2,	                   % sets default tabsize to 2 spaces
   title=\lstname                   % show the filename of files included with \lstinputlisting; also try caption instead of title
}

@@

In this section, we evaluate the proposed matrix-free algorithms for finite-strain hyperelastic materials as well as the geometric multigrid preconditioner on a benchmark problem from \cite{Miehe2007}. The coarse meshes for the 2D and 3D problems are illustrated in Figure \ref{fig:miehe}. The 2D material consists of a square (depicted in blue), a hole and two spherical inclusions (depicted in red). The size of the square domain is $10^{-3} \rm{mm}$.
The matrix material is taken to have Poisson's ratio $0.3$ and shear modulus $\mu=0.4225 \times 10^6\, \rm{N/mm^2}$. The inclusion is taken to be $100$ times stiffer.
The 3D material is obtained by extrusion of the 2D geometry into the third dimension.
Note that although a relatively coarse mesh is used at the coarsest multigrid level, the geometry of the inclusions is captured accurately by manifold descriptions of the boundaries and interfaces.
Both domains are fully fixed along the bottom surface and a distributed load is applied at the top in the $(1,0)$ or $(1,1,0)$ directions for the 2D and the 3D problems, respectively.
The force density $12.5 \times 10^3 \rm{N/mm^2}$ or $12.5 \sqrt{2} \times 10^3 \rm{N/mm^3}$ is applied in 5 steps for the 2D and the 3D problems, respectively.
With respect to the nonlinear solver, the displacement tolerance of the $\mathcal{\ell}_2$ norm for the Newton update is taken to be $10^{-5}$ whereas the relative and absolute tolerances for the residual forces is $10^{-8}$. The relative convergence criteria for the linear solver is $10^{-6}$.
%
Figure \ref{fig:miehe_deformed} shows deformed meshes at the final loading step with quadratic elements and two global mesh refinements.
%
\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.35\textwidth}
      \centering
      \includegraphics[width=\textwidth]{deformed.png}
      \caption{2D deformed mesh}
  \end{subfigure}
  \begin{subfigure}[b]{0.35\textwidth}
    \centering
    \includegraphics[width=\textwidth]{deformed_3d.png}
    \caption{3D deformed mesh}
  \end{subfigure}
  \caption{Deformed meshes at the final loading step with quadratic elements and two global mesh refinements.}%
  \label{fig:miehe_deformed}
\end{figure}
%
\begin{table}[!htb]
  \caption{Parameters for the benchmark: $p$ is the polynomial degree,
  $q$ is the number of quadrature points in 1D, $N_{\text{gref}}$ is the number of global mesh refinements, $N_{el}$ is the number of elements and $N_{\text{DoF}}$ is the number of DoFs.}
  \begin{subtable}{.5\linewidth}
  \caption{2D}
  \label{tab:input_parameters_2d}
  \centering
  \begin{tabular}{ccccc}
  \hline
    $p$ & $q$ & $N_{\text{gref}}$ & $N_{el}$ & $N_{\text{DoF}}$ \\
  \hline
    1 & 2 & 7 & 1441792 & 2887680 \\
    2 & 3 & 6 & 360448 & 2887680 \\
    3 & 4 & 5 & 90112 & 1625088 \\
    4 & 5 & 5 & 90112 & 2887680 \\
    5 & 6 & 5 & 90112 & 4510720 \\
    6 & 7 & 4 & 22528 & 1625088 \\
    7 & 8 & 4 & 22528 & 2211328 \\
    8 & 9 & 4 & 22528 & 2887680 \\
  \hline
  \end{tabular}
  \end{subtable}
  \begin{subtable}{.5\linewidth}
  \caption{3D}
  \label{tab:input_parameters_3d}
  \centering
  \begin{tabular}{ccccc}
  \hline
    $p$ & $q$ & $N_{\text{gref}}$ & $N_{el}$ & $N_{\text{DoF}}$ \\
  \hline
    1 & 2 & 4 & 1441792 & 4442880 \\
    2 & 3 & 3 & 180224 & 4442880 \\
    3 & 4 & 2 & 22528 & 1891008 \\
    4 & 5 & 2 & 22528 & 4442880 \\
  \hline
  \end{tabular}
  \end{subtable}
\end{table}

@@

\begin{figure}[!ht]
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{CSL_Munich_cg2d.eps}
    \caption{CG iterations (2D)}
    \label{fig:benchmark_miehe_Emmy_cg2}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{CSL_Munich_cg3d.eps}
    \caption{CG iterations (3D)}
    \label{fig:benchmark_miehe_Emmy_cg3}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{CSL_Munich_solver2d.eps}
    \caption{CG solution time (2D)}
    \label{fig:benchmark_miehe_Emmy_sol2}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{CSL_Munich_solver3d.eps}
    \caption{CG solution time (3D)}
    \label{fig:benchmark_miehe_Emmy_sol3}
  \end{subfigure}
  \caption{Cascade Lake, one socket with 20 cores, iterative solver.}%
  \label{fig:benchmark_miehe_Emmy_cg}
\end{figure}

\begin{figure}[!ht]
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{IWR_newest_patched_cg2d.eps}
    \caption{CG iterations (2D)}
    \label{fig:benchmark_miehe_IWR_cg2}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{IWR_newest_patched_cg3d.eps}
    \caption{CG iterations (3D)}
    \label{fig:benchmark_miehe_IWR_cg3}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{IWR_newest_patched_solver2d.eps}
    \caption{CG solution time (2D)}
    \label{fig:benchmark_miehe_IWR_sol2}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{IWR_newest_patched_solver3d.eps}
    \caption{CG solution time (3D)}
    \label{fig:benchmark_miehe_IWR_sol3}
  \end{subfigure}
  \caption{IWR cluster, 20 cores.}%
  \label{fig:benchmark_miehe_IWR_cg}
\end{figure}

Next, we evaluate the performance of the proposed geometric multigrid preconditioner.
Our main goal is to examine whether or not the adopted level operators lead to an efficient preconditioner from the linear algebra perspective. To that end,
we consider the average number of CG iterations throughout the entire simulation, i.e., averaged over each Newton-Raphson iteration and each loading step.
Figures \ref{fig:benchmark_miehe_Emmy_cg2}, \ref{fig:benchmark_miehe_Emmy_cg3}, \ref{fig:benchmark_miehe_IWR_cg2} and \ref{fig:benchmark_miehe_IWR_cg3} show a
 noticeable increase in the average number of CG iterations for increasing polynomial degrees
 for the GMG preconditioner.
Although not reported in this section, we identified that
the coarse level solver setup will greatly affect the performance of the preconditioner.
In particular, we can improve the number of iterations by choosing a more accurate coarse level solver.
The black-box AMG preconditioner
demonstrates the same trend in terms of the number of CG iterations, but
requires many more iterations for convergence.
Consequently, we can conclude that (i) the selected smoother is suitable for the present finite strain elasticity problem, and that (ii) although the adopted level operators are not built using the triple product $\gz A^{l+1}=\gz I^{l+1}_{l} \gz A^l \gz I^l_{l+1}$, they still result in a multigrid preconditioner which requires fewer CG iterations as compared to the algebraic multigrid baseline.

>>>>

\usepackage{pgfplots}
\usepackage{bm}
\usepackage{xcolor}
\usepackage{listings}
\lstset{ %
   backgroundcolor=\color{white},   % choose the background color; you must add \usepackage{color} or \usepackage{xcolor}
   basicstyle=\small,        % the size of the fonts that are used for the code
   breakatwhitespace=false,         % sets if automatic breaks should only happen at whitespace
   breaklines=false,                 % sets automatic line breaking
   captionpos=b,                    % sets the caption-position to bottom
   %commentstyle=\color{mygreen},    % comment style
   deletekeywords={...},            % if you want to delete keywords from the given language
   escapeinside={\%*}{*)},          % if you want to add LaTeX within your code
   extendedchars=false,              % lets you use non-ASCII characters; for 8-bits encodings only, does not work with UTF-8
   frame=single,	                   % adds a frame around the code
   keepspaces=true,                 % keeps spaces in text, useful for keeping indentation of code (possibly needs columns=flexible)
   keywordstyle=\color{blue},       % keyword style
   language=C++,                 % the language of the code
   otherkeywords={*,...},           % if you want to add more keywords to the set
   numbers=left,                    % where to put the line-numbers; possible values are (none, left, right)
   numbersep=5pt,                   % how far the line-numbers are from the code
   numberstyle=\tiny\color{gray}, % the style that is used for the line-numbers
   rulecolor=\color{black},         % if not set, the frame-color may be changed on line-breaks within not-black text (e.g. comments (green here))
   showspaces=false,                % show spaces everywhere adding particular underscores; it overrides 'showstringspaces'
   showstringspaces=false,          % underline spaces within strings only
   showtabs=false,                  % show tabs within strings adding particular underscores
   stepnumber=2,                    % the step between two line-numbers. If it is 1, each line will be numbered
%   stringstyle=\color{mymauve},     % string literal style
   tabsize=2,	                   % sets default tabsize to 2 spaces
   title=\lstname                   % show the filename of files included with \lstinputlisting; also try caption instead of title
}

@@

In this section, we evaluate the proposed matrix-free algorithms for finite-strain hyperelastic materials as well as the geometric multigrid preconditioner on a benchmark problem from \cite{Miehe2007}. The coarse meshes for the 2D and 3D problems are illustrated in Figure \ref{fig:miehe}. The 2D material consists of a square (depicted in blue), a hole and two spherical inclusions (depicted in red). The size of the square domain is $10^{-3} \rm{mm}$.
The matrix material is taken to have Poisson's ratio $0.3$ and shear modulus $\mu=0.4225 \times 10^6\, \rm{N/mm^2}$. The inclusion is taken to be $100$ times stiffer.
The 3D material is obtained by extrusion of the 2D geometry into the third dimension.
Note that although a relatively coarse mesh is used at the coarsest multigrid level, the geometry of the inclusions is captured accurately by manifold descriptions of the boundaries and interfaces.
Both domains are fully fixed along the bottom surface and a distributed load is applied at the top in the $(1,0)$ or $(1,1,0)$ directions for the 2D and the 3D problems, respectively.
The force density $12.5 \times 10^3 \rm{N/mm^2}$ or $12.5 \sqrt{2} \times 10^3 \rm{N/mm^3}$ is applied in 5 steps for the 2D and the 3D problems, respectively.
With respect to the nonlinear solver, the displacement tolerance of the $\mathcal{\ell}_2$ norm for the Newton update is taken to be $10^{-5}$ whereas the relative and absolute tolerances for the residual forces is $10^{-8}$. The relative convergence criteria for the linear solver is $10^{-6}$.
%
Figure \ref{fig:miehe_deformed} shows deformed meshes at the final loading step with quadratic elements and two global mesh refinements.
%
\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.35\textwidth}
      \centering
      \includegraphics[width=\textwidth]{deformed.png}
      \caption{2D deformed mesh}
  \end{subfigure}
  \begin{subfigure}[b]{0.35\textwidth}
    \centering
    \includegraphics[width=\textwidth]{deformed_3d.png}
    \caption{3D deformed mesh}
  \end{subfigure}
  \caption{Deformed meshes at the final loading step with quadratic elements and two global mesh refinements.}%
  \label{fig:miehe_deformed}
\end{figure}
%
\begin{table}[!htb]
  \caption{Parameters for the benchmark: $p$ is the polynomial degree,
  $q$ is the number of quadrature points in 1D, $N_{\text{gref}}$ is the number of global mesh refinements, $N_{el}$ is the number of elements and $N_{\text{DoF}}$ is the number of DoFs.}
  \begin{subtable}{.49\linewidth}
  \caption{2D}
  \label{tab:input_parameters_2d}
  \centering
  \begin{tabular}{ccccc}
  \hline
    $p$ & $q$ & $N_{\text{gref}}$ & $N_{el}$ & $N_{\text{DoF}}$ \\
  \hline
    1 & 2 & 7 & 1441792 & 2887680 \\
    2 & 3 & 6 & 360448 & 2887680 \\
    3 & 4 & 5 & 90112 & 1625088 \\
    4 & 5 & 5 & 90112 & 2887680 \\
    5 & 6 & 5 & 90112 & 4510720 \\
    6 & 7 & 4 & 22528 & 1625088 \\
    7 & 8 & 4 & 22528 & 2211328 \\
    8 & 9 & 4 & 22528 & 2887680 \\
  \hline
  \end{tabular}
  \end{subtable}
  \begin{subtable}{.49\linewidth}
  \caption{3D}
  \label{tab:input_parameters_3d}
  \centering
  \begin{tabular}{ccccc}
  \hline
    $p$ & $q$ & $N_{\text{gref}}$ & $N_{el}$ & $N_{\text{DoF}}$ \\
  \hline
    1 & 2 & 4 & 1441792 & 4442880 \\
    2 & 3 & 3 & 180224 & 4442880 \\
    3 & 4 & 2 & 22528 & 1891008 \\
    4 & 5 & 2 & 22528 & 4442880 \\
  \hline
  \end{tabular}
  \end{subtable}
\end{table}

@@

\begin{figure}[!ht]
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{CSL_Munich_cg2d.eps}
    \caption{CG iterations (2D)}
    \label{fig:benchmark_miehe_CSL_cg2}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{CSL_Munich_cg3d.eps}
    \caption{CG iterations (3D)}
    \label{fig:benchmark_miehe_CSL_cg3}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{CSL_Munich_solver2d.eps}
    \caption{CG solution time (2D)}
    \label{fig:benchmark_miehe_Emmy_sol2}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{CSL_Munich_solver3d.eps}
    \caption{CG solution time (3D)}
    \label{fig:benchmark_miehe_Emmy_sol3}
  \end{subfigure}
  \caption{Cascade Lake, one socket with 20 cores, iterative solver.}%
  \label{fig:benchmark_miehe_Emmy_cg}
\end{figure}

\begin{figure}[!ht]
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{IWR_newest_patched_cg2d.eps}
    \caption{CG iterations (2D)}
    \label{fig:benchmark_miehe_IWR_cg2}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{IWR_newest_patched_cg3d.eps}
    \caption{CG iterations (3D)}
    \label{fig:benchmark_miehe_IWR_cg3}
  \end{subfigure}
  ~
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{IWR_newest_patched_solver2d.eps}
    \caption{CG solution time (2D)}
    \label{fig:benchmark_miehe_IWR_sol2}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \centering
    \includegraphics[width=\textwidth]{IWR_newest_patched_solver3d.eps}
    \caption{CG solution time (3D)}
    \label{fig:benchmark_miehe_IWR_sol3}
  \end{subfigure}
  \caption{IWR cluster, 20 cores.}%
  \label{fig:benchmark_miehe_IWR_cg}
\end{figure}

Next, we evaluate the performance of the proposed geometric multigrid preconditioner.
Our main goal is to examine whether or not the adopted level operators lead to an efficient preconditioner from the linear algebra perspective. To that end,
we consider the average number of CG iterations throughout the entire simulation, i.e., averaged over each Newton-Raphson iteration and each loading step.
Figures \ref{fig:benchmark_miehe_CSL_cg2}, \ref{fig:benchmark_miehe_CSL_cg3}, \ref{fig:benchmark_miehe_IWR_cg2} and \ref{fig:benchmark_miehe_IWR_cg3} show a
 noticeable increase in the average number of CG iterations for increasing polynomial degrees
 for the GMG preconditioner.
The black-box AMG preconditioner
demonstrates the same trend in terms of the number of CG iterations, but
requires many more iterations for convergence.
Consequently, we can conclude that (i) the selected smoother is suitable for the present finite strain elasticity problem, and that (ii) although the adopted level operators are not built using the triple product $\gz A^{l+1}=\gz I^{l+1}_{l} \gz A^l \gz I^l_{l+1}$, they still result in a multigrid preconditioner which requires fewer CG iterations as compared to the algebraic multigrid baseline.
